---
title: "Human Development Indicators Analysis"
format:
  revealjs:
    theme: serif
    slideNumber: true
    css: custom.css
---

## Title

Human Development Indicators Analysis

A Comparative Study of Singapore and Switzerland

Using data.table in

Presented by: Jane Sanjeevini

## Introduction

In this presentation, we will:

-   Introduce the Human Development Indicators (HDI) dataset.

-   Perform data cleaning and merging using data.table.

-   Explore and analyze the HDI data for Singapore and Switzerland.

-   Visualize key findings using R.

## Data Description

The HDI dataset includes:

-   `country_code`: ISO country code

-   `country_name`: Country name

-   `indicator_id`: Indicator ID

-   `indicator_name`: Indicator name

-   `index_id`: Index ID

-   `index_name`: Index name

-   `value`: Indicator value

-   `year`: Year of observation

## Data Preparation

-   Data loading and preparation using data.table

-   Reading CSV files for Singapore and Switzerland

-   Assigning correct data types to variables

## Loading The Data

**singapore \<- fread("hdro_indicators_sgp.csv")**

**switzerland \<- fread("hdro_indicators_che.csv")**

```{r}
library(data.table)

# Load the Singapore and Switzerland data
singapore <- fread("hdro_indicators_sgp.csv")
switzerland <- fread("hdro_indicators_che.csv")

# Display the structure and first few rows of the datasets
str(singapore)
str(switzerland)

head(singapore)
head(switzerland)

```

## Removing Missing Values 

**singapore \<- singapore\[!is.na(as.numeric(value)), \]**

**switzerland \<- switzerland\[!is.na(as.numeric(value)), \]**

**singapore\[, `:=` (value = as.numeric(value), year = as.integer(year))\]**

**switzerland\[, `:=` (value = as.numeric(value), year = as.integer(year))\]**

**singapore \<- na.omit(singapore)switzerland \<- na.omit(switzerland)**

## Appropriation of Data

Convert columns to appropriate data types and handle missing values.

```{r}
# Identify non-numeric values in the 'value' column for Singapore
non_numeric_singapore <- singapore[!is.na(as.numeric(value)) == FALSE]

# Identify non-numeric values in the 'value' column for Switzerland
non_numeric_switzerland <- switzerland[!is.na(as.numeric(value)) == FALSE]

# Display the problematic values
print(non_numeric_singapore)
print(non_numeric_switzerland)

# Remove non-numeric values from Singapore dataset
singapore <- singapore[!is.na(as.numeric(value)), ]

# Remove non-numeric values from Switzerland dataset
switzerland <- switzerland[!is.na(as.numeric(value)), ]

# Convert columns to appropriate data types after cleaning
singapore[, `:=` (value = as.numeric(value), year = as.integer(year))]
switzerland[, `:=` (value = as.numeric(value), year = as.integer(year))]

# Handle missing values
singapore <- na.omit(singapore)
switzerland <- na.omit(switzerland)

```

## Merging the Datasets

-   **Merging Datasets**: Combined datasets from Singapore and Switzerland using `rbindlist`, creating a unified dataset for comparative analysis of human development indicators.

-   **Ensuring Data Consistency**: Added a 'country' column to distinguish between Singapore and Switzerland data post-merge, facilitating clear identification and analysis of country-specific trends.

    **combined_data \<- rbindlist(list(singapore, switzerland))**

```{r}
# Add a country column
singapore[, country := "Singapore"]
switzerland[, country := "Switzerland"]

# Merge the datasets
combined_data <- rbindlist(list(singapore, switzerland))

```

## Data Cleaning

-   **Handling Missing Data**: Identified and removed non-numeric values in critical columns like 'value', ensuring dataset integrity and accuracy for analysis.

-   **Standardizing Data Types**: Converted variables to appropriate data types (numeric for 'value', integer for 'year'), ensuring consistency and compatibility across datasets from different sources.

```{r}
# Remove negative and excessively high values from 'value' column
combined_data <- combined_data[value >= 0 & value <= 1000]   

# Check for remaining non-numeric values in 'value' column
non_numeric_values <- combined_data[is.na(as.numeric(value))]
print(non_numeric_values)  

# Convert 'value' to numeric after cleaning
combined_data[, value := as.numeric(value)]

```

## Summary

**unique_indicators \<- unique(combined_data\$indicator_name)**

**unique_years \<- unique(combined_data\$year)**

```{r}
# Summary of cleaned data
summary(combined_data)

# Check unique indicators and years again
unique_indicators <- unique(combined_data$indicator_name)
unique_years <- unique(combined_data$year)

print(unique_indicators)
print(unique_years)

```

## Data Exploration

-   The dataset comprises 1588 observations of human development indicators across countries from 1990 to 2022.

-   Numeric values for indicators range widely (0.752 to 97.614), offering insights into diverse metrics over time.

-   Key focus areas include life expectancy, education attainment, inequality measures, environmental impact, and gender equality, enabling comprehensive global trend analysis.

## Key Indicators Over Time

Analyzing and plotting the key indicators over time for both countries.

**life_expectancy \<- combined_data\[indicator_name == "Life Expectancy at Birth (years)", .(mean_value = mean(value)), keyby = .(country, year)\]**

```{r}
library(data.table)
library(ggplot2)

# Filter data for Life Expectancy at Birth
life_expectancy <- combined_data[indicator_name == "Life Expectancy at Birth (years)",
                                 .(mean_value = mean(value)),
                                 keyby = .(country, year)]

# Define custom colors for Life Expectancy at Birth plot
custom_colors <- c("Singapore" = "plum",
                   "Switzerland" = "darkviolet")

# Plot: Life Expectancy at Birth Over Time
plot_life_expectancy <- ggplot(life_expectancy, aes(x = year, y = mean_value, color = country)) +
  geom_line() +
  scale_color_manual(values = custom_colors) +
  labs(title = "Life Expectancy at Birth Over Time", x = "Year", y = "Life Expectancy (years)") +
  theme_minimal()

# Display the plot
plot_life_expectancy
```

## Education Index Analysis

Analyzing and plotting the education index for Singapore and Switzerland.

**mean_schooling \<- combined_data\[indicator_name == "Mean Years of Schooling (years)", .(mean_value = mean(value)), keyby = .(country, year)\]**

```{r}
# Filter data for Mean Years of Schooling
mean_schooling <- combined_data[indicator_name == "Mean Years of Schooling (years)",
                                .(mean_value = mean(value)),
                                keyby = .(country, year)]

# Define custom colors for Mean Years of Schooling plot
custom_colors <- c("Singapore" = "violet",
                   "Switzerland" = "salmon")

# Plot: Mean Years of Schooling Over Time
plot_mean_schooling <- ggplot(mean_schooling, aes(x = year, y = mean_value, color = country)) +
  geom_line() +
  scale_color_manual(values = custom_colors) +
  labs(title = "Mean Years of Schooling Over Time", x = "Year", y = "Mean Years of Schooling") +
  theme_minimal()

# Display the plot
plot_mean_schooling

```

## Carbon Dioxide Emissions 

Analyzing and plotting the emission of carbon dioxide for the two countries.

**co2_emissions \<- combined_data\[indicator_name == "Carbon dioxide emissions per capita (production) (tonnes)", .(mean_value = mean(value)), keyby = .(country, year)\]**

```{r}
# Filter data for CO2 emissions per capita
co2_emissions <- combined_data[indicator_name == "Carbon dioxide emissions per capita (production) (tonnes)",
                               .(mean_value = mean(value)),
                               keyby = .(country, year)]

# Define custom colors for Singapore and Switzerland
custom_colors <- c("Singapore" = "purple",  
                   "Switzerland" = "pink")

# Plot: CO2 Emissions per Capita Over Time
plot_co2_emissions <- ggplot(co2_emissions, aes(x = year, y = mean_value, color = country)) +
  geom_line() +
  scale_color_manual(values = custom_colors) +
  labs(title = "CO2 Emissions per Capita Over Time", x = "Year", y = "CO2 Emissions (tonnes)") +
  theme_minimal()

# Display the plot
plot_co2_emissions

```

## Interpretation

-   Life expectancy has generally increased over the years, though the rate of increase has slowed in recent times.

-   The average number of years of schooling has been rising steadily, indicating improvements in education levels over time.

-   CO2 emissions per capita have shown an upward trend, indicating increasing environmental impact per person over the years.

## Conclusion

-   **Life Expectancy**: Improved due to healthcare and living conditions in countries like the US, Singapore, and Switzerland, indicating significant public health advancements.

-   **Educational Attainment**: Mean Years of Schooling have increased steadily, supporting personal and economic growth, with ongoing data clarity challenges.

-   **Environmental Impact**: Rising CO2 Emissions per Capita underscore sustainability challenges despite human development gains, necessitating stronger environmental policies in nations like Singapore and Switzerland.

## Q&A

1.  How has life expectancy at birth changed over time across different countries?
2.  What is the relationship between education attainment (mean years of schooling) and socio-economic factors like income inequality and labor force participation rates?
3.  How have carbon dioxide emissions per capita evolved over the past decades, and how do they correlate with indicators of human development and environmental sustainability?

## References

The references or resources used in the presentation are:

Singapore Dataset: "<https://data.humdata.org/dataset/hdro-data-for-singapore>"

Switzerland Dataset: "<https://data.humdata.org/dataset/hdro-data-for-switzerland>"
